---
title: "Team 1 Stepping Stones Variables: Final"
author: "Fara Islam, Lily Slonim, Lekha Mereddy"
date: "2023-03-02"
output:   
    html_document:
      toc: true
      toc_depth: 4
      toc_float: true
      code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = FALSE)
```

```{r, show_col_types = FALSE}
# Load libraries
library(tidyverse)
library(scales) #control plot scales
library(ggthemes) #color pallets and customized themes
library(janitor)

# Reading in/prepping data ----

# Pre-collected variables: Math SOL Pass Rate for Grade 3, 5, and 8
math_sol_pass <- read_csv("../data/sol_pass.csv") %>% filter(subject=="Mathematics")

math_sol_pass <- math_sol_pass %>%
  mutate(division_name = str_remove(division_name, "County|City"),
         division_name = str_trim(division_name),
         year = str_sub(school_year, 6,9))

miss2021 <- data.frame(division_name = c("Albemarle", "Charlottesville", "Virginia"),
                       year = c("2020", "2020", "2020"))

## Grade 3
math_sol_pass_grade3 <- math_sol_pass %>% filter(test_level=="Grade 3") %>% 
                        bind_rows(miss2021)

## Grade 5
math_sol_pass_grade5 <- math_sol_pass %>% filter(test_level=="Grade 5") %>% 
                        bind_rows(miss2021)

## Grade 8 
math_sol_pass_grade8 <- math_sol_pass %>% filter(test_level=="Grade 8") %>% 
                        bind_rows(miss2021)


# Variable 1: Economically Disadvantaged Students
economically_dis <- read_csv("../data/economically_disadvantaged_students.csv")

economically_dis <- economically_dis %>%
  mutate(division_name = str_remove(division_name, "County|City"),
         division_name = str_trim(division_name),
         year = str_sub(school_year, 6,9))

# Variable 2: Education-Residents Over the Age 25 with High School Diploma/Equivalent or Higher (%)
hsdegree_attainment <- read_csv("../data/hsdegree_attainment.csv") 

hsdegree_attainment <- hsdegree_attainment %>% 
  mutate(NAME = str_remove(NAME, " County, Virginia| city, Virginia"), 
         NAME = str_trim(NAME))

# Variable 3: School Reports of Physical Violence

physical_violence <- read_csv("../data/school_physical_violence.csv") 
  
# Variable 4: Children in Foster Care
foster_care <- read_csv("../data/foster_care.csv")

# Variable 5: Homeless adults (CSV File only)
homeless_count <- read_csv("../data/pitcount_homeless.csv") 
```

## Introduction 
Stepping Stones was created to chart the health of children and their families in the City of Charlottesville and Albemarle County. Stepping Stones includes data on 38 indicators chosen to measure child and family well-being. Stepping Stones is intended to provide accurate data, stimulate community dialogue and action regarding regional trends in the Charlottesville/Albemarle community.



## Education 

### Math SOL Pass Rate - Grade 3 {.tabset} 

#### Trend
```{r Math SOL - Grade 3 Graph}

math_sol_pass_grade3 %>% 
ggplot(aes(x = year, y = pass_rate, color = division_name,group = division_name)) +
  geom_line() +
  geom_point() + 
  theme_minimal() +
  geom_vline(xintercept = "2020", linetype = 2, color = "black") +
  annotate("text", x = 11.5, y = 95, size = 2, hjust = 0, color = "black", 
           label = 
"On March 13, The Virginia 
Governor ordered that all of 
Virginia’s K-12 schools close 
for the COVID-19 Pandemic. Thus, 
SOLs were not conducted in 2020") +
  geom_vline(xintercept = "2012", linetype = 2, color = "black") +
  annotate("text", x =4.2, y = 10, size = 2, hjust = 0, color = "black", 
           label = 
"The Virginia Department 
of Education releases a 
new version of the Math SOL") +
  scale_color_manual(values = c("#B23B6B", "#E7A342", "#5AB48A")) +
  scale_y_continuous(limits = c(0, 100),
                     labels = label_percent(scale = 1), 
                     name = "Percent") + 
  scale_x_discrete(name = "Year") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position="top") + 
  labs(title = "Academic Achievement of Grade 3 Students",
       subtitle = "The percent of grade 3 students who pass the math SOL",
       caption = "Source: Virginia Department of Education (VDOE). 2006 - 2021. 
       https://p1pe.doe.virginia.gov/buildatable/testresults", 
       color = "") 


```

#### About
The Virginia Standards of Learning (SOLs) are a set of educational standards that outline what students should know and be able to do at each grade level in the state of Virginia. The SOLs cover a wide range of subjects, including English, mathematics, science, history and social science, foreign languages, health, physical education, and the fine arts. Pass rate reports are used to assess student achievement, evaluate school and district performance, and determine whether schools are meeting state and federal standards. The Virginia Department of Education (VDOE) reports pass rates for each SOL assessment for all public schools in Virginia. See [VDOE](https://p1pe.doe.virginia.gov/buildatable/testresults) for more detail.

* **Why is this variable important?** The [Standards of Learning (SOL)](https://www.doe.virginia.gov/teaching-learning-assessment/student-assessment/virginia-sol-assessment-program) tests are a requirement for Virginia public schools. The SOL establishes a minimum expectation of students in various subjects. [Third grade is a relevant year to examine because the curriculum focuses on educational fundamentals](https://www.aecf.org/resources/early-warning-why-reading-by-the-end-of-third-grade-matters#findings-and-stats). The proceeding grades build upon these skills. Students without a firm grasp of these fundamentals can be more likely to [drop out of school in later years](https://www.aecf.org/resources/early-warning-why-reading-by-the-end-of-third-grade-matters#findings-and-stats). Math is a fundamental subject that impacts the everyday lives of people. Additionally, the Math SOL provides a way for Charlottesville, Albemarle County, and Virginia to understand academic success in STEM-related fields by students.

* **Further Considerations:** Examining the third-grade math SOL pass rate by socioeconomic status, race, and ethnicity could emphasize specific trends within demographic groups. This could provide more direction as to which populations of students would benefit from early interventions from school staff. 

* **Trend:** Charlottesville’s third-grade math SOL pass rate was higher than Virginia and Albemarle County between 2011 - 2015. Between 2016 - 2017, Charlottesville's pass rate was higher than Albemarle County's but lower than the state average. Since 2016, Charlottesville's pass rate has been under Virginia and Albemarle County. The pass rate was relatively high from 2006 - 2011, however, the pass rate dropped in 2012 across Charlottesville, Albemarle County, and Virginia. The sharp decline is likely attributed to an [updated version of the Math SOL](https://www.wtvr.com/2012/08/14/virginia-sol-test-scores-school), which enhanced the overall rigor of the test. In 2020, the pass rate declined across all localities because schools were closed during the [COVID-19 Pandemic](http://charlottesvilleschools.org/coronavirus-archived-page/). While the 2021 pass rate across localities has increased since 2020, the pass rate is still below the pre-pandemic level. 

* **How do we measure this?** The pass rate of students is count of 3rd grade students who passed the Math SOL over the count of all 3rd grade students in the school division that took the Math SOL.

* **Data Source:** Virginia Department of Education (VDOE). 2006 - 2021. https://p1pe.doe.virginia.gov/buildatable/testresults

* **Files on Github:** https://github.com/virginiaequitycenter/stepping-stones/blob/main/code/sol_pass.R

### Math SOL Pass Rate - Grade 5 {.tabset} 

#### Trend
```{r Math SOL - Grade 5 Graph}

math_sol_pass_grade5 %>% 
ggplot(aes(x = year, y = pass_rate, color = division_name,group = division_name)) +
  geom_line() +
  geom_point() + 
  theme_minimal() +
  geom_vline(xintercept = "2020", linetype = 2, color = "black") +
  annotate("text", x = 11.5, y = 95, size = 2, hjust = 0, color = "black", 
           label = 
"On March 13, The Virginia 
Governor ordered that all of 
Virginia’s K-12 schools close 
for the COVID-19 Pandemic. Thus, 
SOLs were not conducted in 2020") +
  geom_vline(xintercept = "2012", linetype = 2, color = "black") +
  annotate("text", x =4.2, y = 10, size = 2, hjust = 0, color = "black", 
           label = 
"The Virginia Department 
of Education releases a 
new version of the Math SOL") +
  scale_color_manual(values = c("#B23B6B", "#E7A342", "#5AB48A")) +
  scale_y_continuous(limits = c(0, 100),
                     labels = label_percent(scale = 1), 
                     name = "Percent") + 
  scale_x_discrete(name = "Year") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position="top") + 
  labs(title = "Academic Achievement of Grade 5 Students",
       subtitle = "The percent of grade 5 students who pass the math SOL",
       caption = "Source: Virginia Department of Education (VDOE). 2006 - 2021. 
       https://p1pe.doe.virginia.gov/buildatable/testresults", 
       color = "")


```

#### About
The Virginia Standards of Learning (SOLs) are a set of educational standards that outline what students should know and be able to do at each grade level in the state of Virginia. The SOLs cover a wide range of subjects, including English, mathematics, science, history and social science, foreign languages, health, physical education, and the fine arts. Pass rate reports are used to assess student achievement, evaluate school and district performance, and determine whether schools are meeting state and federal standards. The Virginia Department of Education (VDOE) reports pass rates for each SOL assessment for all public schools in Virginia. See [VDOE](https://p1pe.doe.virginia.gov/buildatable/testresults) for more detail.

* **Why is this variable important?** The [Standards of Learning (SOL)](https://www.doe.virginia.gov/teaching-learning-assessment/student-assessment/virginia-sol-assessment-program) tests are a requirement for Virginia public schools. The SOL establishes a minimum expectation of students in various subjects. The math SOL pass rate provides a quantitative measure to understand the effectiveness of the math education program for students in fifth grade. [Fifth grade is an important year to study because students are solidifying skills learned in lower grades](https://assets.aecf.org/m/resourcedoc/AECF-Early_Warning_Full_Report-2010.pdf). Math is a fundamental subject that impacts the everyday lives of people. Additionally, the Math SOL provides a way for Charlottesville, Albemarle County, and Virginia to understand academic success in STEM-related fields by students. It is also important to examine pass rates in relationship to the COVID-19 Pandemic. Albemarle county and Charlottesville have not yet rebounded from the [achievement gap](https://www.nbc29.com/2022/08/18/sol-test-results-are-released-2021-2022/) that started during the pandemic. This gap is most notable in schools that were [closed for in-person learning](https://www.nbc29.com/2022/08/18/sol-test-results-are-released-2021-2022/). 

* **Further Considerations:** Charlottesville has had a lower fifth-grade math SOL pass rate than Virginia and Albemarle County since 2006, with 2007 being the only exception. Examing Charlottesville’s fifth-grade math curriculum and resources compared to Virginia and Albemarle County could provide insight into the pass rate trend. Additionally, analyzing the fifth-grade math SOL pass rate by socioeconomic status, race, and ethnicity could emphasize specific trends within demographic groups. This could provide more direction as to which populations of students would benefit from early interventions from school staff.

* **Trend:** Charlottesville has had a lower fifth-grade math SOL pass rate than Virginia and Albemarle County since 2006, with 2007 being the only exception. The Math SOL pass rate was relatively high from 2006 - 2011, however, the pass rate dropped in 2012 across Charlottesville, Albemarle County, and Virginia. The sharp decline is likely attributed to an [updated version of the Math SOL](https://www.wtvr.com/2012/08/14/virginia-sol-test-scores-school), which enhanced the overall rigor of the test. In 2020, the pass rate declined across all localities because schools were closed during the [COVID-19 Pandemic](http://charlottesvilleschools.org/coronavirus-archived-page/). While the 2021 pass rate across localities has increased since 2020, the pass rate is still below the pre-pandemic level. 

* **How do we measure this?** The pass rate of students is count of 5th grade students who passed the Math SOL over the count of all 5th grade students in the school division that took the Math SOL.

* **Data Source:** Virginia Department of Education (VDOE). 2006 - 2021. https://p1pe.doe.virginia.gov/buildatable/testresults

* **Files on Github:** https://github.com/virginiaequitycenter/stepping-stones/blob/main/code/sol_pass.R

### Math SOL Pass Rate - Grade 8 {.tabset} 

#### Trend
```{r Math SOL - Grade 8 Graph}

math_sol_pass_grade8 %>% 
ggplot(aes(x = year, y = pass_rate, color = division_name, group = division_name)) +
  geom_line() +
  geom_point() + 
  theme_minimal() +
  geom_vline(xintercept = "2020", linetype = 2, color = "black") +
  annotate("text", x = 11.5, y = 95, size = 2, hjust = 0, color = "black", 
           label = 
"On March 13, The Virginia 
Governor ordered that all of 
Virginia’s K-12 schools close 
for the COVID-19 Pandemic. Thus, 
SOLs were not conducted in 2020") +
  geom_vline(xintercept = "2012", linetype = 2, color = "black") +
  annotate("text", x =4.2, y = 10, size = 2, hjust = 0, color = "black", 
           label = 
"The Virginia Department 
of Education releases a 
new version of the Math SOL") +
  scale_color_manual(values = c("#B23B6B", "#E7A342", "#5AB48A")) +
  scale_y_continuous(limits = c(0, 100),
                     labels = label_percent(scale = 1), 
                     name = "Percent") + 
  scale_x_discrete(name = "Year") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position="top") + 
  labs(title = "Academic Achievement of Grade 8 Students", 
       subtitle = "The percent of grade 8 students who pass the math SOL",
       caption = "Source: Virginia Department of Education (VDOE). 2006 - 2021. 
       https://p1pe.doe.virginia.gov/buildatable/testresults", 
       color = "")


```

#### About
The Virginia Standards of Learning (SOLs) are a set of educational standards that outline what students should know and be able to do at each grade level in the state of Virginia. The SOLs cover a wide range of subjects, including English, mathematics, science, history and social science, foreign languages, health, physical education, and the fine arts. Pass rate reports are used to assess student achievement, evaluate school and district performance, and determine whether schools are meeting state and federal standards. The Virginia Department of Education (VDOE) reports pass rates for each SOL assessment for all public schools in Virginia. See [VDOE](https://p1pe.doe.virginia.gov/buildatable/testresults) for more detail.

* **Why is this variable important?** The [Standards of Learning (SOL)](https://www.doe.virginia.gov/teaching-learning-assessment/student-assessment/virginia-sol-assessment-program) tests are a requirement for Virginia public schools. The SOL establishes a minimum expectation of students in various subjects. The math SOL pass rate provides a quantitative measure to understand the effectiveness of the math education program for students in eighth grade. Math is a fundamental subject that impacts the everyday lives of people. Additionally, the Math SOL provides a way for Charlottesville, Albemarle County, and Virginia to understand academic success in STEM-related fields by students. It is important to examine pass rates in relationship to the COVID-19 Pandemic. Albemarle county and Charlottesville have not yet rebounded from the [achievement gap](https://www.nbc29.com/2022/08/18/sol-test-results-are-released-2021-2022/) that started during the pandemic. This gap is most notable in schools that were [closed for in-person learning](https://www.nbc29.com/2022/08/18/sol-test-results-are-released-2021-2022/). 

* **Further Considerations:** Examining the eighth grade math SOL pass rate by socioeconomic status, race, and ethnicity could emphasize specific trends within demographic groups. This could provide more direction as to which populations of students would benefit from early interventions from school staff. 

* **Trend:** In Virginia, the Grade 8 Math SOL pass rate has been relatively consistent over the years. The pass rate has ranged from a low of 76.1% in 2015 to a high of 81.8% in 2019, with an average pass rate of 79.1%. For Charlottesville, the Grade 8 Math SOL pass rate has been consistently below the state average over the years. The pass rate has ranged from a low of 63.4% in 2015 to a high of 70.2% in 2019, with an average pass rate of 66.9%. For Albemarle, the Grade 8 Math SOL pass rate has been consistently above the state average over the years. The pass rate has ranged from a low of 78.1% in 2016 to a high of 86.6% in 2017, with an average pass rate of 82.6%. Over the years, the pass rate for Albemarle has been consistently higher than the pass rate for Charlottesville. The difference in pass rates between Albemarle and Charlottesville ranges from a low of 7.1% in 2017 to a high of 23.6% in 2016. After 2019, it can be noticed that there is a drastic decline in pass rates across all three domains. It is safe to assume that the pandemic led to the closure of schools and the shift to online learning, which may have disrupted the learning process for many students. The sudden transition to online learning and the associated challenges, such as a lack of access to technology, limited opportunities for collaboration, and difficulties in adjusting to a new learning environment, could have negatively impacted student performance on SOL exams.

* **How do we measure this?** The pass rate of students is count of 8th grade students who passed the Math SOL over the count of all 8th grade students in the school division that took the Math SOL.

* **Data source:** Virginia Department of Education, Test Results Build-A-Table, 2006-2021.https://p1pe.doe.virginia.gov/buildatable/testresults

* **Files on Github:** https://github.com/virginiaequitycenter/stepping-stones/blob/main/code/sol_pass.R


### Highschool Degree Attainment {.tabset}
#### Trend

```{r}

# hsdegree_attainment %>% 
#   ggplot(aes(x = year, y = pcthshigherE, color = NAME,group = NAME)) +
#   geom_line() +
#   geom_point()+
# #  geom_vline(xintercept = 2021, linetype = 2, color = "black") +
#   # annotate("text", x = 4.4, y = 10, size = 2, hjust = 0, color = "black",
#   #          label =
#   #          "State Transfer Module legislation
#   #          passed in 2004 required that two‐
#   #          and four‐year institutions develop
#   #          dual admissions agreements and programs,
#   #          reflected in the increasing trend in Virginia")
#   theme_minimal()+
#   theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position="top") +
#   scale_x_continuous(breaks = seq(2010, 2021, 1))+
#   scale_y_continuous(limits = c(80, 100),
#                      labels = label_percent(scale = 1), 
#                      name = "Percent") + 
#   scale_x_discrete(name = "Year") +
#   scale_color_manual(values = c("#B23B6B", "#E7A342", "#5AB48A")) +
#   labs(title = "Education attainment of residents over the age of 25", 
#        subtitle =  "The percent of residents with a High School Diploma/Equivalent or greater", 
#        caption = "Source: Virginia Department of Education",
#        color="") 

  

hsdegree_attainment %>%
  ggplot(aes(x = year, y = pcthshigherE, color = NAME,group = NAME)) +
  geom_line() +
  geom_point()+
  theme_minimal()+
 geom_vline(xintercept = 2021, linetype = 2, color = "black") +
  annotate("text", x = 2018.8, y = 85, size = 2, hjust = 0, color = "black",
         label ="State Transfer Module legislation\npassed in 2004, which required\nthat two and four year\ninstitutions develop dual\nadmissions agreements\nand programs, possibly\ncaused the increasing\ntrend in Virginia in later years") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position="top", legend.title = element_blank()) +
  scale_x_continuous(breaks = seq(2010, 2021, 1), name = "Year")+
  scale_y_continuous(limits = c(80, 95),
                     labels = label_percent(scale = 1),
                     name = "Percent") + 
  scale_color_manual(values = c("#B23B6B", "#E7A342", "#5AB48A")) +
  labs(title = "Education Level of Residents Over the Age of 25", subtitle =  "The percent of residents 25+ with a high school diploma/equivalent or greater", caption = "Source: U.S. Census Bureau, American Community Survey 5-year estimates,\n 'Educational Attainment (S1501).' 2010 - 2021.\n https://data.census.gov/table?q=S1501&tid=ACSST1Y2021.S1501") 
  


```

#### About
This section looks at the education levels of residents Over the Age 25 with High School Diploma/Equivalent or Higher. The data is collected from [Census Data](https://data.census.gov/table?tid=ACSST1Y2021.S1501).

* **Why is this variable important?:** The numbers and trends on the graph could be representative of resources and funding on a city, county, and state level and whether or not they differ, and how this can be improved to make equal opportunities available to more people, to contribute to better education levels. Earnings and employment in this population can be related with education level data to understand if and how education levels contribute to a better standard of living. Education policy can be improved as this data can aid policymakers in understanding what kinds of resources are available in these areas and ultimately push for reforms to improve education levels and funding. Higher education levels may contribute to greater productivity, growth and contribution to society which in turn benefits each individual as well.

* **Further Considerations:** It may also be significant for immigration and economic policy in the areas. Education rates may be affected by data of those who have not been educated in the area (so local resource application may not apply to them), but the outcomes can be understood to better delegate opportunities.

* **Trend:** When comparing the graduation rates to the percentage of students enrolled in post-secondary education, there is a large number of students who are not continuing their education. However, trends show that there are instances of greater secondary education enrollment, for example in 2018. Additionally, in 2004, Virginia passed the [State Transfer Model legislation](https://lis.virginia.gov/cgi-bin/legp604.exe?041+ful+CHAP0553) which aimed to facilitate transfers to public and private four-year institutions of higher education. This data looks at individuals above the age of 25, which means many of them were in high school around the time this legislation was implemented, and could contribute to the overall growing trend of the data.

* **How do we measure this?** The percentage of residents over the age of 25 with high school diplomas or more is calculated by dividing it by the total population in the city, county and state.
  
* **Data Source:** U.S. Census Bureau, American Community Survey 5-year estimates, "Educational Attainment (S1501)." 2010 - 2021. https://data.census.gov/table?q=S1501&tid=ACSST1Y2021.S1501

* **Files on Github:**https://github.com/virginiaequitycenter/stepping-stones/blob/main/code/hsdegree_attainment.R


## Discipline in Schools

### School Reports of Physical Violence {.tabset} 

#### Trend

```{r}

physical_violence %>% 
ggplot(aes(x = year, y = rate, color = division_name, group = division_name)) +
  geom_line() +
  geom_point() + 
  theme_minimal() +
  geom_vline(xintercept = 2020, linetype = 2, color = "black") +
  annotate("text", x = 2019.42, y = 95, size = 2, hjust = 0, color = "black", 
           label = 
"On March 13, The Virginia 
Governor ordered that all of 
Virginia’s K-12 schools close 
for the COVID-19 Pandemic") +
  geom_vline(xintercept = 2020, linetype = 2, color = "black") +
  annotate("text", x =2020.05 , y = 50, size = 2, hjust = 0, color = "black",
           label = "In 2020,Albemarle\nCounty Public Schools\nadopts the\nSTEP program") +
  scale_color_manual(values = c("#B23B6B", "#E7A342", "#5AB48A")) +
  scale_y_continuous(limits = c(0, 100),
                     labels = label_percent(scale = 1), 
                     name = "Reports of School Violance Rate (per 1000 students)") + 
  scale_x_continuous(name = "Year") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position="top") + 
  labs(title = "School Reports of Physical Violence",
       subtitle = "The rate of school reports of physical violence per 1000 students",
       caption = "Source: Virginia Department of Education (VDOE). Discipline, Crime & Violence. 2018 - 2021.
       https://www.doe.virginia.gov/data-policy-funding/data-reports/data-collection/special-education", 
       color = "") 


```


#### About 

* **Why is this variable important?** School reports of physical violence can be an indicator of student health. Physical violence can cause [physical and emotional harm to students](https://www.cdc.gov/violenceprevention/pdf/school_violence_fact_sheet-a.pdf). These experiences can lead to [negative health behaviors,  including alcohol and drug use](https://www.cdc.gov/violenceprevention/pdf/school_violence_fact_sheet-a.pdf). Reports of physical violence can lead to school discipline. Understanding a potential cause that leads to school discipline is important because there is a relationship between [school discipline and graduation rates, arrests, and incarceration](https://www.gse.harvard.edu/news/uk/19/09/school-discipline-linked-later-consequences). Examining school reports of physical violence concerning the COVID-19 Pandemic is also valuable. Nationally, school teachers, personnel, and administrators experienced an [increase in harassment, both verbal and physical](https://www.apa.org/news/press/releases/2022/03/school-staff-violence-pandemic), from students since the 2020 Pandemic. 

* **Further Considerations:** Examining the reports of physical violence by socioeconomic status, race, and ethnicity could emphasize specific trends within demographic groups. 

* **Trend:** Virginia and Albemarle’s reports of school violence rate have been declining since 2018. Charlottesville had an increase in the reports of school violence rate, which aligns with the national trend. There is a sharp decline for all localities in 2021. In 2020, Albemarle County Public Schools adopted [STEP](https://www.nbc29.com/2020/02/07/albemarle-county-public-schools-program-provides-alternative-out-of-school-suspension/), a program to reduce out-of-school suspensions. This could be one cause of the decline in 2021 for Albemarle County. 

* **How do we measure this?** School reports of physical violence are created by adding the incidents in various incident categories (assault battery, fighting conflict, sexual offenses, threats, kidnapping, and robbery). The total count of physical violence reports is divided by the number of students enrolled in the school district or state. Lastly, this number is multiplied by 1000.    

* **Data Source:** Virginia Department of Education (VDOE). Discipline, Crime & Violence. 2018 - 2021. https://www.doe.virginia.gov/data-policy-funding/data-reports/data-collection/special-education

* **Files on Github:** https://github.com/virginiaequitycenter/stepping-stones/blob/main/code/school_physical_violence.R

## Finances 

### Students Identified as Economically Disadvantaged {.tabset}
#### Trend

```{r Economically Disadvantaged Graph}
economically_dis %>% 
  mutate(percent = (percent/100)) %>% 
  ggplot(aes(x = year, y = percent, color = division_name, group= division_name)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  geom_vline(xintercept = "2020", linetype = 2, color = "black") +
  annotate("text", x = 16.2, y = .55, size = 2.3, hjust = 0, 
           label = "In 2020, in Charlottesville,\nstudents were automatically\nenrolled in the meal\nprogram.") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position="top") +
  scale_y_continuous(labels = scales::percent) + 
  scale_color_manual(values = c("#B23B6B", "#E7A342", "#5AB48A")) +
  labs(title = "Students who are Economically Disadvantaged", 
       subtitle = "The percent of finance-children qualifying for free reduced meals",
       caption = "Source: Virginia Department of Education, Fall Membership Build-A-Table. 2004 - 2023.
       https://p1pe.doe.virginia.gov/buildatable/fallmembership", 
       x = "Year", y = "Percent", color = "")
```


#### About
The National School Lunch Program (NSLP) provides free or reduced-price lunches to students from low-income families. The eligibility criteria are based on household income and family size. Schools that participate in the NSLP receive federal funding to help cover the cost of the meals. This metric is used to determine whether a student is economically disadvantaged. See [VDOE](https://p1pe.doe.virginia.gov/buildatable/testresults) for more detail.

* **Why is this important?** These programs help to ensure that students from low-income families have access to nutritious meals, which can support their health, growth, and academic success. By promoting equal access to education, improving student health, supporting families in need, and encouraging academic success, the provision of free and reduced-price meals is an important part of efforts to promote equity and improve outcomes for all students.

* **Further Considerations:** Looking at the percentage of children in that are economically disadvantaged by race and ethnicity could also highlight important trends.

* **Trend:** In 2020, the percentage of economically disadvantaged students in Virginia was 46.8%, while the percentages for Charlottesville and Albemarle were 56.4% and 28.4%, respectively. This suggests that Charlottesville has a higher concentration of economically disadvantaged students compared to the state average, while Albemarle has a lower concentration. The data shows that the percentage of economically disadvantaged students in Virginia has been increasing over the past few years, although at a relatively slow rate. In 2016, the percentage was 44.6%, while in 2020, it was 46.8%. The percentage of economically disadvantaged students in both Charlottesville and Albemarle has been increasing over the past few years. In 2020, the percentage in Charlottesville was 56.4%, up from 48.7% in 2016, while in Albemarle, it was 28.4%, up from 21.9% in 2016. 

* **How do we measure this?** The percent of students is count of students identified as eligible free reduced lunch over the count of all students in the school division.

* **Data source:** Virginia Department of Education, "Fall Membership Build-A-Table." 2004-2023. https://p1pe.doe.virginia.gov/buildatable/fallmembership

* **Files on Github:** https://github.com/virginiaequitycenter/stepping-stones/blob/main/code/economically_disadvanaged.R


## Family Characteristics 

### Children in Foster Care {.tabset}
#### Trend

```{r}
foster_care %>% 
  #mutate(fc_rate = (fc_rate/100)) %>% 
  ggplot(aes(x = year, y = fc_rate, color = locality)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  geom_vline(xintercept = 2020, linetype = 2, color = "black") +
  annotate("text", x = 2020.15, y = .16, size = 2.5, hjust = 0, 
           label = "The state’s 2020\nbudget for foster care\nwas greatly reduced\nin response to\nCOVID-19.") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  #scale_y_continuous(labels = scales::percent) + 
  scale_x_continuous(breaks = 2006:2022) +
  scale_color_manual(values = c("#B23B6B", "#E7A342", "#5AB48A")) +
  theme(legend.position="top") + 
  labs(title = "Children in Foster Care", 
       subtitle = "The rate of children in foster care (per 1000 children)",
       caption = "Source: Virginia Department of Social Services, 
       Foster Care Reports, Children Demographics. 2006-2022.
       https://www.dss.virginia.gov/geninfo/reports/children/fc.cgi", 
       x = "Years", y = "Children in Foster Care (per 1000 children)", color = "")

```

#### About
The Virginia Department of Social Services (VDSS) publishes a range of reports on the foster care system in the state. These reports provide information on the number and demographics of children in foster care, the types of foster care placements, and the outcomes of children who have been in foster care. See [VDSS](https://www.dss.virginia.gov/geninfo/reports/children/fc.cgi) for more detail.

* **Why is this important?** Understanding the foster care system in Virginia and the experiences of children in foster care is important for several reasons. Firstly, it helps protect the well-being of children who have often experienced abuse, neglect, or other traumatic events. Secondly, it promotes equity by identifying and addressing disparities among different racial or ethnic groups. Thirdly, it can help improve outcomes by ensuring that children have the best possible chance for a successful future. Finally, data and information on the foster care system in Virginia can help policymakers and practitioners identify trends and areas for improvement, which can be used to inform policies and practices and create a more effective and equitable foster care system.

* **Further Considerations:** Analyzing the percentage of children in foster care by US Census Tract could highlight spatial patterns. Additionally, looking at the percentage of children in foster care by socio-economic status, race and ethnicity could also highlight important trends. 

* **Trend:** The percentage of children in foster care in Virginia has steadily increased over the past four years. In 2016, there were 4,429 children in foster care, while in 2020 there were 5,604 children in foster care, representing an increase of approximately 27%. In Charlottesville, the number of children in foster care has fluctuated but has generally increased from 2016 to 2020. In 2016, 82 children were in foster care, while in 2020, 96 children were in foster care. In Albemarle County, the number of children in foster care has steadily increased from 2016 to 2020. In 2016, there were 46 children in foster care, while in 2020 there were 73 children in foster care, representing an increase of approximately 58%. The rate of foster care placements is higher in Charlottesville than in Albemarle County. In 2020, the rate of foster care placements in Charlottesville was 7.3%, while in Albemarle County, the rate was 4.8%. There is a decrease in the percentage of children in foster care starting in 2018. This is likely caused by programs aimed to reduce separation of families. There is uncertainty around if this is the same narrative post-pandemic. 

* **How do we measure this?** The percent of students is count of students identified as foster children over the count of all students in the school division.

* **Data source:** Virginia Department of Social Services, Foster Care Reports, "Children Demographics." 2006-2022. https://www.dss.virginia.gov/geninfo/reports/children/fc.cgi

* **Files on Github:** https://github.com/virginiaequitycenter/stepping-stones/blob/main/code/children_fostercare.R


## Housing

### Adults experiencing Homelessness {.tabset}
#### Trend


```{r }
homeless_count %>% 
  filter(type == "total_persons") %>% 
 ggplot(aes(x = year, y = count)) +
   geom_line(color = "#E7A342") +
   geom_point(color = "#E7A342") + 
   theme_minimal() +
  geom_vline(xintercept = 2022, linetype = 2, color = "black") +
  annotate("text", x = 2021 , y = 150, size = 2.5, hjust = 0.7, 
           label = "The pandemic has\nbrought in high housing\ncosts and lack of available\naffordable housing") +
  scale_y_continuous(limits = c(100,300)) + 
  scale_x_discrete(name = "Year") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Adults Experiencing Homelessness",
        subtitle = "The total number of adults experising homeless in the City of Charlottesville",
        caption = "Source: Blue Ridge Area Coalition for the Homeless (BRACH)",
        y = "Total Persons") 

```


#### About
The Blue Ridge Area Coalition for the Homelesss (BRACH) aims to make homelessness rare, brief and nonrecurring in our community. Some of the key services offered by BRACH include emergency shelter, transitional housing, case management, and job training and placement assistance. The organization also provides food, clothing, and other basic necessities to those in need. BRACH relies on the support of volunteers and donations from individuals, businesses, and organizations to fund its programs and services. Its work has made a significant impact in the community, helping to provide crucial support to those who are struggling with homelessness and working towards a future where everyone has access to safe and stable housing.

* **Why is this important?** Knowing the population of homelessness is crucial to identify the needs of the homeless population and allocate resources effectively. Accurate data on the homeless population can also help set specific goals, collaborate with other organizations, and advocate for policy changes. By using data to inform its work, the project can make a positive impact in the lives of homeless individuals and families in the community. Additionally, by monitoring the population of homelessness, it is possible to evaluate the effectiveness of interventions and programs, and make necessary adjustments to ensure that resources are being used effectively. 

* **Further Considerations:** Homelessness can be affected by seasonal factors, like weather conditions and changes in the economy. Therefore along with looking at trends over multiple years to identify patterns, further data can be considered which looks at seasonal fluctuations, which may more accurately identify the causes and policy effects of this data. Changes in the demographics of the homeless population, such as changes in the age or gender of those experiencing homelessness, can also provide insight into the underlying causes of homelessness and inform targeted interventions. Trends in the homeless population can also be affected by changes in the availability and affordability of housing in Charlottesville. Examining trends in housing prices, evictions, and rental vacancies can provide context for changes in the homeless population. Changes in the size and demographics of the homeless population can also be influenced by changes in outreach and intervention efforts, such as increased funding for homeless services or changes in how services are delivered. It's important to examine how changes in these efforts may be impacting the homeless population over time. Finally, looking at this data in comparison with nearby cities, counties, and the state as a whole may also be useful in better understanding trends and opportunities for policy changes.

* **Trend:** The trend over the years shows general stability until 2022 where there is a rise in the number. This could be attributed to the COVID 19 pandemic that occured over the previous two years, which brought [high housing costs and a lack of available housing](https://infocville.com/2022/05/03/council-briefed-on-increase-in-homelessness-in-charlottesville-area/#:~:text=%E2%80%9CThe%20most%20significant%20shift%20has,affordable%20housing%2C%E2%80%9D%20Haro%20said.)

* **How do we measure this?** The total number of persons experiencing homeless in Charlottesville. This was the only data obtained from BRACH. 

* **Data source:** Blue Ridge Area Coalition for the Homeless(BRACH)

